The Impact of Alternation

### 2017-05-01 draft.tex
first version

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\Sigma}$ consists of two distinct sorts
$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
\in \ssorts{\sig}$. If $n=0$, $f$ is a \emph{constant} of sort
$\sigma(f)$ and we consider the constants $\top,\bot \in \sfuns{\sig}$
of sort $\Bool$, denoting the truth values true and false.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $S$ is a variable $x \in
\vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$.  A formula is in
disjunctive normal form (DNF) if it is a disjunction of conjunctions
of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
\item the $\Bool$ sort into the set $\Bool^\I$ containing the boolean
  values, and
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}

>>>>

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\Sigma}$ consists of two distinct sorts
$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
\in \ssorts{\sig}$. If $n=0$, $f$ is a \emph{constant} of sort
$\sigma(f)$. We consider a constant $\mathbf{0}$ of sort $\Data$ and
$\top,\bot$ of sort $\Bool$, denoting the truth values true and false.

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $S$ is a variable $x \in
\vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$.  A formula is in
disjunctive normal form (DNF) if it is a disjunction of conjunctions
of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$. We
denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
and write $\fv{}{\phi}$ for $\bigcup_{S \in \ssorts{\sig}}
\fv{S}{\phi}$.

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
\item the $\Bool$ sort into the set $\Bool^\I$ containing the boolean
  values, and
\item each function symbol $f$ into a total function $f^\I :
  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$. We omit
mentioning the interpretation when it is clear from the context.

\section{Alternating Data Automata}

In the rest of this section we fix an interpretation $\I$ and a finite
alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
\subset \vars$ of variables of sort $\Data$, a \emph{data word} is a
finite sequence of pairs $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$
where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : \vec{x}
\rightarrow \Data^\I$ are valuations. We denote by $\vec{x} \mapsto
\Data^\I$ the set of valuations of the variables $\vec{x}$ and by
$\left(\Sigma \times (\vec{x} \mapsto \Data^\I)\right)^*$ the set of
data words over $\vec{x}$.

Given two sets of variables $B,X \subset \vars$, we denote by
$\Form^+(B,X)$ the set of formulae $\phi$ such that $\fv{\Bool}{\phi}
\subseteq B$, $\fv{\Data}{\phi} \subseteq X$ and each variable from
$B$ occurs under an even number of negations.

An \emph{alternating data automaton} (ADA) is a tuple $\A =
\tuple{\vec{x},Q,I,F,\Delta}$, where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $I,F \subseteq Q$ are sets of \emph{initial} and \emph{final}
  states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
current values of the variables of $\A$. Formally, let $\vec{x}_i =
\set{x_i \mid x \in \vec{x}}$, $i\geq0$, be a time-stamped set of
variables. For an input even $a \in \Sigma$ and a formula
$\phi(\vec{x}_0,\ldots,\vec{x}_i)$ we write $\Delta(\phi,a)$ for the
formula obtained from $\phi$ by simultaneously replacing each state $q
\in \fv{\Bool}{\phi}$ by the formula
$\Delta(q,a)[\vec{x}_i/\overline{\vec{x}},\vec{x}_{i+1}/\vec{x}]$, for
all $i\geq0$.

Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$, the
\emph{run} of $\A$ over $w$ is the sequence of formulae
$\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \bigwedge_{q \in
  I} q$ and, for all $i > 0$, we have $\phi_i \equiv
\Delta(\phi_{i-1},a_i)$. We slightly abuse notation and write
$\Delta(\phi_0,a_1,\ldots,a_n)$ for
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. $\A$ 
\emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is the
valuation that maps:\begin{inparaenum}[(1)] 
\item each $x \in \vec{x}_0$ to $\mathbf{0}$, 
%
\item each $x \in \vec{x}_i$ to $\nu_i(x)$, for all $i \in [1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
